---
title: "HBoV-1 Persistence Modeling"
date: "`r format(Sys.time(),'%B, %d, %Y')`"
output:
  rmdformats::material:
    highlight: kate
    code_folding: hide
    fig_caption: true
bibliography: boca_library.bib
---

```{r knitr_init, echo = FALSE, cache = FALSE, warning = FALSE}
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(rmdformats))

## Global options
options(max.print = "75")
opts_chunk$set(echo = TRUE,
	             cache = TRUE,
               prompt = FALSE,
               tidy = FALSE,
               comment = NA,
               message = FALSE,
               warning = FALSE)
opts_knit$set(width = 85)
```

# Introduction

Human bocavirus (HBoV-1), a DNA virus first identified in 2005 [@allander_cloning_2005], has been frequently detected in young children experiencing acute respiratory tract illness. HBoV-1 has been detected in up to 18% of samples from children with respiratory illness [@jartti_human_2012]. Over 85% of children in the United States have antibodies to this virus by 4 years of age [@kahn_seroepidemiology_2008], yet HBoV-1 is rarely detected in adults [@jartti_human_2012]. Furthermore, HBoV-1 is often detected in children without evidence of disease and is often identified alongside with other respiratory viruses associated with acute illness [@jartti_human_2012, @martin_clinical_2008, @fry_human_2007, @allander_human_2007, @christensen_human_2010]. Thus, retrospective or cross-sectional analyses designed to make an inference on disease progression, incidence and persistence are problematic. The purpose of this analysis is in identifying trends regarding persistence and latency of human bocavirus 1 in a pediatric cohort.

## Data Source

Data for the current study comes from previously reported prospective studies of human herpesvirus 6 (HHV-6) natural history and human bocavirus-1 shedding events from a group of infants [@zerr_population-based_2005], [@martin_human_2015]. Incidence of HHV-6 has been reported between 64-83% in the United States and in some studies up to 90% prevalence has been reported [@braun_human_1997]. For this reason, the entire cohort of children carried HHV-6 at one point in the 2-year study period.

## Study Population

Inclusion criteria for enrollment included pregnancy of the mother, receipt of care at a Seattle area-based obstetrical practice, and provision of informed consent prior to participation. Children were then followed from birth for up to two years between April 1997 and August 2003. No further exclusion criteria were applied within the birth cohort.

## Clinical Data Collection

Demographic data were collected from the mother at enrollment through administered surveys. Symptom data were collected using a daily diary recorded by the mother and/or father. Symptoms included: fever (temperature >38.0&deg; C), roseolla cough, runny-nose, vomiting, diarrhea, rash, fussiness above the baseline level for the child, seizure, and physician visit for illness. Data was further collected on breast-feeding, breast-feeding duration, group child-care, daycare attendance and duration and playgroup attendance.

## Methods 

Infection episodes were defined as at least two weeks of RT-PCR positive HBoV-1 infection for entry and exit with no more than 1 interrim negative specimen within the episode. The first episode was termed primary with all others defined as secondary, tertiary, quaternary, and quinary. For purposes of analyses all secondary and subsequent infection episodes were categorized as 'secondary'.

# R Packages Used

```{r library loading}
library(lubridate)    # time and date functions
library(padr)         # pad time series with missing values
library(grid)         # grid.arrange
library(gridExtra)    # grid.arrange extras
library(Rmisc)        # convenience functions for analysis
library(ggthemes)     # theme_tufte, theme_bw, etc.
library(gridExtra)    # arrange ggplot grobs
library(RColorBrewer) # plotting colors
library(lattice)      # versatile plotting functions
library(magrittr)     # extra pipelines/infix operators (%<>%)
library(reshape2)     # melt and cast data, i.e. wide to long, long to wide
library(tidyverse)    # everything, including ggplot, dplyr, etc.
library(knitr)        # kable() tables
library(kableExtra)   # kable options
```

# Load Data

```{r}
setwd("~/Documents/Martin Lab/HBov-1/Persistence_modelling/BocavirusSEATTLE/Rdata")
# setwd("S:/MartinEpi/Analysis/Bocavirus Persistence Modeling/Rdata/")
dem.clin.dat <- readRDS("dem_clin_dat.rds")
diaries <- readRDS("diaries.rds")
diary.pos <- readRDS("diary_pos.rds")
id <- readRDS("pos_id.rds")
main.long <- readRDS("main_long.rds")
main.pos <- readRDS("main_pos.rds")
pcr.by.week <- readRDS("pcr_by_week.rds")
results.dob <- readRDS("results_dob.rds")
season.decomp <- readRDS("season_decomp.rds")
test.results.pos <- readRDS("test_results_pos.rds")
bov.pad <- readRDS("bov_pad.rds")
```

```{r, eval = TRUE, echo = FALSE}
# misc. functions
# true minimal theme
theme_trueMinimal <- function(base_size = 11, base_family = "") {
    theme_bw(base_size = base_size, base_family = base_family) %+replace% 
        theme(axis.ticks = element_blank(), 
              legend.background = element_blank(), 
              legend.key = element_blank(), 
              panel.background = element_blank(), 
              strip.background = element_blank(), 
              panel.border = element_blank(),
              panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              plot.background = element_blank(), 
              axis.line = element_line(), 
              axis.line.y = element_blank(),
              complete = TRUE)
}
```

# Explore Data

First I want to define a persistent infection. Should it be anything greater than 4 weeks? 6 weeks? 10 weeks? I am going to first explore the data visually to get a better sense of the distributions. 

Further, It would not seem fair to examine those with infections between 2 and 10 weeks vs. 10 to $max(time)$ or any split for that matter.

```{r}
infx_grp <- bov.pad %>%
  filter(!(episode %in% c('','blip'))) %>% 
  group_by(id, episode) %>% 
  arrange(id, week) %>% 
  select(-bov_infx, -log_q_per_ml, -primary_cnt)

# infection durations
infx_dur <- infx_grp %>% 
  mutate(time = row_number(), 
         duration = max(time)) %>% 
  ungroup(id) %>% 
  distinct(id, episode, duration) %>% 
  reshape2::dcast(., id ~ episode, value.var = 'duration')
  
colnames(infx_dur)[-1] <- paste(colnames(infx_dur)[-1], "_dur", sep = "")

# grab beginning and end dates of infections
infx_dates <- infx_grp %>% 
  mutate(first = row_number() == 1,
         last = row_number() == n()) %>% 
  filter(!(first == FALSE & last == FALSE)) %>% 
  ungroup()

start <- infx_dates %>% 
  filter(first == TRUE) %>% 
  select(-first, -last) %>% 
  group_by(id) %>% 
  spread(episode, week)

colnames(start)[-1] <- paste(colnames(start)[-1], "_start", sep = "")
  
end <- infx_dates %>% 
  filter(last == TRUE) %>% 
  select(-first, -last) %>% 
  group_by(id) %>% 
  spread(episode, week)

colnames(end)[-1] <- paste(colnames(end)[-1], "_end", sep = "")

# merge into one data.frame of durations
durations <- join_all(list(infx_dur, start, end), 
                      by = "id", type = "full") %>%
  # rearranging for 'tidyness'
  select(id, starts_with("primary"), 
             starts_with("secondary"),
             starts_with("tertiary"),
             starts_with("quaternary"),
             starts_with("quinary"))
```

```{r}
sumFunc <- function(x) {
  n = length(na.omit(x))
  min = min(x, na.rm = TRUE)
  max = max(x, na.rm = TRUE)
  median = median(x, na.rm = TRUE)
  mean = mean(x, na.rm = TRUE)
  sd = sd(x, na.rm = TRUE)
  return(list(n = n,
              mean = mean,
              sd = sd,
              median = median,
              min = min, max = max))
}

summary_table <- sapply(select(durations, ends_with("dur")), sumFunc)
colnames(summary_table) <- sub("_dur", "", colnames(summary_table))
knitr::kable(summary_table)
```

```{r}
my_cols <- RColorBrewer::brewer.pal(9, "Set1")

durations %>% 
  select(ends_with("dur")) %>% 
  melt() %>% 
  filter(!is.na(value)) %>% 
  rename(episode = variable,
         duration = value) %>% 
  mutate(episode = factor(episode,
                           levels = c("primary_dur",
                                      "secondary_dur",
                                      "tertiary_dur",
                                      "quaternary_dur",
                                      "quinary_dur"),
                           labels = sub("_dur", "", 
                                        colnames(summary_table)))) %>% 
  ggplot(aes(x = episode, y = duration)) %>% 
  add(geom_boxplot(color = my_cols[2], 
                   outlier.color = my_cols[5],
                   outlier.size = 3,
                   outlier.alpha = 0.5)) %>% 
  add(geom_jitter(color = my_cols[5], alpha = 0.5, size = 3)) %>% 
  add(theme_trueMinimal()) %>%
  add(labs(x = "Infection Episode",
           y = "Infection Duration, weeks"))
```

There were `r summary_table[1, 1]` primary infections of the 86 children tested for HBoV-1 by RT-PCR. Primary infections have a mean of `r round(as.numeric(summary_table[2, 1]), 2)` weeks of persistent shedding (95% CI: `r round(as.numeric(t.test(durations$primary_dur)$conf.int[1]), 2)`, `r round(as.numeric(t.test(durations$primary_dur)$conf.int[2]), 2)`), with a median of `r summary_table[4, 1]` weeks and maximum of `r summary_table[6, 1]`.

Subsequent shedding episodes generally decrease in duration, with increasing mean durations occurring later in the child's life history (quaternary episode) possibly due to recurrence or end of dormancy. Through genetic testing, our group has come to the conclusion that these multiple periods of infection are likely from the same infecting strain, giving some weight to the idea of an end of latency giving rise to a peak in late shedding.

```{r}
quantile(durations$primary_dur, seq(0, 1, by = 0.1))
```

70% of the infections are less than 6 weeks in duration. We could consider infections of 10 weeks or more as being persistent infections (20% of the data).

## Persistent Infections

```{r}
persistent <- durations %>% 
  filter(primary_dur >= 10) %>%
  select(id, primary_dur)

persistent %>%
  summarise(n = n(),
            median = median(primary_dur),
            iqr = IQR(primary_dur),
            mean = mean(primary_dur),
            sd = sd(primary_dur),
            min = min(primary_dur),
            max = max(primary_dur))
```

Defining persistent shedding events as > 10 weeks results in 14 individuals with a mean duration of 17 weeks (95 % CI: `r round(as.numeric(t.test(persistent$primary_dur)$conf.int[1]), 2)`, `r round(as.numeric(t.test(persistent$primary_dur)$conf.int[2]), 2)`)

```{r}
primary_persist <- persistent %>%
  select(id) %>%
  left_join(bov.pad, by = "id") %>%
  filter(episode == 'primary') %>%
  select(-primary_cnt, -bov_infx, -episode) %>%
  arrange(id, week) %>%
  group_by(id) %>%
  mutate(time = row_number()) %>%
  ungroup()
  
grp <- unique(primary_persist$id)[1:7]
ppA <- primary_persist %>% filter(id %in% grp)
ppB <- primary_persist %>% filter(!(id %in% grp))

grpPlot <- function(df) {
  p <- df %>%
  filter(!is.na(log_q_per_ml)) %>%
  ggplot(aes(x = time)) %>%
  add(geom_line(aes(y = log_q_per_ml, color = id),
                size = 1.05)) %>%
  add(geom_hline(yintercept = 3.0, linetype = 2)) %>%
  add(theme_bw()) %>%
  add(scale_x_continuous(breaks = seq(0, 30, by = 5),
                         limits = c(0, 30))) %>%
  add(scale_y_continuous(breaks = seq(2, 10, by = 2),
                         limits = c(2, 10))) %>%
  add(theme(legend.position = c(0.9, 0.8))) %>%
  add(labs(title = '',
           x = '',
           y = '',
           color = "Infant ID"))
  p
}

grpAplot <- grpPlot(ppA)
grpBplot <- grpPlot(ppB)
```

```{r, fig.cap = "--- indicates lower limit of detection = 3.0", fig.width = 10, fig.height = 7}
grid.arrange(grpAplot, grpBplot, ncol = 2,
             bottom = grid::textGrob("Weeks from Infection Start"),
             left = grid::textGrob(expression(Log[10] ~ HBoV-1/mL), rot = 90))
```

By splitting into groups of 7 the shapes of the infectivity curves are a little clearer, though not by much. At a cursory glance, some shedding patterns tend to start low, increase to a peak, then decay exponentially with some resurgence at the end. On the other hand, some start high and immediately begin to decay exponentially, either one could be true. Also, it is possible that, for the latter case, the study team missed the first positive in the series. We will refer to the first pattern (start low, peak, decay) as pattern 1, and the second one as pattern 2 (start high, decay).

To get a better view of the individual shedding experience I will plot each experience and fit a smoothed line.

```{r, fig.width = 12, fig.height = 7}
primary_persist %>%
  filter(!is.na(log_q_per_ml)) %>%
  ggplot(aes(x = time)) %>%
  add(geom_point(aes(y = log_q_per_ml), color = my_cols[5])) %>%
  add(geom_smooth(aes(y = log_q_per_ml),
                  color = my_cols[2],
                  alpha = 0.1,
                  se = FALSE)) %>%
  add(facet_wrap(~id, ncol = 7)) %>%
  add(theme_trueMinimal()) %>%
  add(scale_x_continuous(breaks = seq(0, 30, by = 5),
                         limits = c(0, 30))) %>%
  add(labs(x = "\n Weeks from Infection Start",
           y = expression(Log[10] ~ HBoV-1/mL),
           caption = "Limited to infections spanning \u2265 10 weeks"))
```

Infants 179, 184, 207, 243, 247, 270, 285, and 317 follow trend 1, while 92, 162, 189, 216, 233, and 271 follow trend 2. Curious as to why that is.

```{r, echo = FALSE, eval = TRUE}
pattern1 <- c('179','184','207','243','247','270','285','317')
pattern2 <- c('92','162','189','216','233','271')
```

This may not matter but I wanted to try my hand at comparing infection curves between primary, secondary, and so on. There are many ways to do this but I am going to use a 'joy plot', a visualization technique that has gained popularity recently that looks pretty dang cool. You can find more information [here](http://blog.revolutionanalytics.com/2017/07/joyplots.html) and [here](http://derekogle.com/fishR/2017-07-28-JoyPlot).

```{r}
library(ggjoy)
bov.pad %>%
  filter(bov_infx == 1 & episode != 'blip') %>%
  arrange(id, week) %>%
  group_by(id) %>%
  mutate(time = row_number()) %>%
  ungroup() %>%
  select(-id, -week, -primary_cnt, -bov_infx) %>%
  filter(!is.na(log_q_per_ml))

```

# References
